\section{Related Work}

The TSP is one of the most intensively studied combinatorial optimization problems.
Its NP-hard nature \cite{garey2002computers} makes exact solution methods infeasible for large instances,
yet the search for optimal solutions has driven significant progress in both exact and heuristic algorithms for solving the TSP
but also in the area of combinatorial optimization as a whole.


\subsection{Exact Solvers}
Exact solutions to the TSP can be obtained by modelling it as a dynamic programming or integer linear programming problem and then solving it using
branch-and-bound or branch-and-cut methods \cite{suiSurveyDeepLearningbased2025}.
The gold standard for exact solvers of the TSP is Concorde \cite{applegateCertificationOptimalTSP2009}, it has been developed and optimized over years specifically to solve TSPs
and was able to obtain an optimal tour for a problem with 85900 cities, computing it required over 136 CPU years \cite{applegateCertificationOptimalTSP2009}.
While exact solvers are the gold standard for verifying solution quality, their runtime increases exponentially, limiting their applicability to large instances.

\subsection{Heuristics and Metaheuristics for the TSP}

To overcome the limitations of exact methods, researchers have developed a wide range of heuristic and metaheuristic algorithms that can find near-optimal solutions in reasonable time.

Heuristics for the TSP can be separated into two groups of algorithms: \textit{tour construction procedures}, which incrementally build a tour by adding a new node to the tour at each step, and \textit{tour improvement procedures}, which start from an initial tour and then try to improve that tour by making changes. \textit{Composite procedures} are algorithms, that start with a solution generated by a construction procedure and then improve it with an improvement procedure \cite{regoLocalSearchMetaheuristics}.

\subsection{History and Developements}

Among the most influential metaheuristics, \textit{Simulated Annealing} (SA) introduced the idea of probabilistic acceptance of worsening moves to escape local optima.

\textit{Tabu Search} (TS), originally proposed by Glover \cite{regoLocalSearchMetaheuristics} and later refined by Gendreau and Potvin \cite{gendreauTabuSearch2019},
uses explicit memory structures (tabu lists) to prevent cycling and guide the search.

\textit{Iterated Local Search} (ILS) \cite{lourencoIteratedLocalSearch2019} builds chains of local optima by perturbing solutions between applications of a local search,
achieving state of the art performance on many problems.

\textit{Genetic Algorithms} (GAs) for the TSP are reviewed by Larra{\~n}aga et al.
\cite{larranagaGeneticAlgorithmsTravelling1999}, and more recent developments,
including the EAX crossover that can find optimal tours for instances with up to 200,000 cities, are discussed by Whitley \cite{whitleyNextGenerationGenetic2019}.

A more in-depth review of metaheuristics and heuristics applied to TSP is provided by Gendreau and Potvin \cite{gendreauHandbookMetaheuristics2019}, where they present a wide variety of both basic and advanced algorithms.

