\section{Materials and Methods}
In this chapter several algorithms are introduced that are at the basis of many state of the art methods for solving the TSP.
The goal is to introduce the reader to a variety of different techniques and concepts that are commonly used.


\subsection{2-Opt}
2-Opt is one of the simplest and most widely used local search heuristics for the TSP \cite{regoLocalSearchMetaheuristics}.
Starting from some initial tour, it repeatedly removes two non-adjacent edges and reconnects the resulting two paths in the only other possible way, which is the same as reversing the segment between the two cut points.
A move is accepted whenever it strictly decreases the tour length.
The neighborhood of a tour with \( n \) cities therefore contains \( \mathcal{O}(n^2) \) candidate moves.
The search terminates in a local minimum, that is, a tour for which no single 2-Opt swap gives an improvement.

In this study, the initial tour is constructed by a deterministic nearest-neighbor heuristic (seeded for reproducibility): starting from a fixed city, the next city is always chosen as the closest unvisited one until all cities are included.
The 2-Opt search itself uses a \emph{first-improvement} strategy: as soon as an improving swap \( (i,j) \) is found, it is applied immediately and the scan is restarted from the beginning of the tour.
This is repeated until a full pass over all pairs \( (i,j) \) finds no improving move, at which point the tour is 2-optimal.

\begin{algorithm}[H]
	\caption{2-Opt}
	\label{alg:two-opt}
	\begin{algorithmic}[1]
		\STATE $s = \text{GreedyInitialTour}()$
		\STATE $\textit{improved} \leftarrow \textbf{true}$
		\WHILE{$\textit{improved}$}
		\STATE $\textit{improved} \leftarrow \textbf{false}$
		\FOR{$i = 0$ \TO $n-3$}
		\FOR{$j = i+2$ \TO $n-1$}
		\STATE $a, b \leftarrow s_i, s_{i+1}$
		\STATE $c, d \leftarrow s_j, s_{(j+1) \bmod n}$
		\IF{$d(a,c) + d(b,d) < d(a,b) + d(c,d)$}
		\STATE reverse $s_{i+1 \ldots j}$
		\STATE $\textit{improved} \leftarrow \textbf{true}$
		\STATE \textbf{break out of both loops}
		\ENDIF
		\ENDFOR
		\ENDFOR
		\ENDWHILE
		\RETURN $s$
	\end{algorithmic}
\end{algorithm}

The improvement condition compares the cost of the two edges that would be removed, \( d(a,b) + d(c,d) \), with the cost of the two edges that would replace them, \( d(a,c) + d(b,d) \).
Because the segment between the two cut points is simply reversed, no other edge weights change and the cost difference can be evaluated in constant time per candidate move.
2-Opt is used here both as a single solver and as the embedded \textit{LocalSearch} of the Iterated Local Search algorithm described below.





\subsection{Simulated Annealing}
Plain local search like 2-Opt always stops at the first local minimum it reaches, because it never accepts a move that makes the tour worse.
Simulated Annealing fixes this by sometimes accepting worse tours on purpose, so that the search can escape a local minimum and keep looking for a better solution \cite{regoLocalSearchMetaheuristics}.

How willing the algorithm is to accept a worse tour is controlled by a single number called \textit{temperature} \( T \).
At each step a neighbor tour is created with one random 2-Opt move, and the change in tour length \( \Delta = \text{Cost}(\text{next}) - \text{Cost}(\text{current}) \) is computed.
A better tour (\( \Delta < 0 \)) is always accepted.
A worse tour is accepted only with probability \( e^{-\Delta / T} \).
When \( T \) is high this probability is close to one, so many worse tours are accepted. When \( T \) is low it is close to zero, so the algorithm behaves almost like normal 2-Opt local search.

The temperature is lowered after every step with a geometric cooling schedule, \( T(t) = T_0 \cdot 0.9999^{\,t} \).
The starting temperature \( T_0 \) is not fixed but estimated for each problem so that the algorithm fits the scale of the distances:
a few hundred random 2-Opt moves are sampled, the average of the worsening moves \( \overline{\Delta} \) (mean delta) is taken, and \( T_0 \) is set so that about \( 60\% \) of such moves would be accepted at the start, i.e. \( T_0 = \overline{\Delta} / \ln(1 / 0.6) \).
The search starts from a greedy nearest-neighbor tour and runs for a fixed budget of \( 100\,000 \) iterations (or stops early once the temperature is practically zero), always returning the best tour seen so far.

\begin{algorithm}[H]
	\caption{Simulated Annealing}
	\label{alg:sa}
	\begin{algorithmic}[1]
		\STATE $T_0 = \text{EstimateInitialTemperature}()$
		\STATE $\textit{current} = \text{GreedyInitialTour}()$
		\STATE $\textit{best} = \textit{current}$
		\FOR{$t = 1$ \TO $\text{MaxIterations}$}
		\STATE $T = T_0 \cdot 0.9999^{\,t}$
		\IF{$T \approx 0$}
		\RETURN $\textit{best}$
		\ENDIF
		\STATE $\textit{next} = \text{Random2OptMove}(\textit{current})$
		\STATE $\Delta = \text{Cost}(\textit{next}) - \text{Cost}(\textit{current})$
		\IF{$\Delta < 0$}
		\STATE $\textit{current} = \textit{next}$
		\IF{$\text{Cost}(\textit{current}) < \text{Cost}(\textit{best})$}
		\STATE $\textit{best} = \textit{current}$
		\ENDIF
		\ELSE
		\STATE accept $\textit{current} = \textit{next}$ with probability $e^{-\Delta / T}$
		\ENDIF
		\ENDFOR
		\RETURN $\textit{best}$
	\end{algorithmic}
\end{algorithm}

A random 2-Opt move is the same edge swap as in normal 2-Opt, but here the two cut points are chosen at random instead of by scanning for an improvement.
This keeps each step cheap while still letting the temperature decide whether a worse tour is kept or thrown away.






\subsection{Tabu Search}
Tabu Search (TS) extends classical local search by using memory structures to escape local minima and to avoid cycling \cite{gendreauTabuSearch2019}.
While a plain local search like 2-Opt stops at the first local optimum, 
TS continues from there by accepting non-improving moves.
To prevent the search from immediately undoing such a move and returning to the previous solution, 
it keeps a \emph{tabu list} of recently performed moves. 
Moves that are \emph{tabu} (forbidden) are not allowed unless they satisfy an \emph{aspiration criterion} (e.g., they lead to a new global best).

In the algorithm implementation \ref{alg:tabu}, the search space is the set of all Hamiltonian tours.
The neighborhood is defined by all 2-Opt swaps, giving $\mathcal{O}(n^2)$ candidate moves per iteration.
The initial solution is constructed by the same greedy nearest-neighbor heuristic used for 2-Opt and Simulated Annealing.

At each iteration, the algorithm examines the whole 2-Opt neighborhood of the current tour.
For every pair $(i,j)$, it generates a neighbor by reversing the segment $i+1 \ldots j$ and computes its total length.
A move is considered admissible if either it is not tabu, or its tour length is better than the best known solution (aspiration criterion).
Among all admissible moves, the one with the smallest tour length is selected (\emph{best improvement} strategy).

\begin{algorithm}[H]
	\caption{Tabu Search}
	\label{alg:tabu}
	\begin{algorithmic}[1]
		\STATE $s \gets \text{GreedyInitialTour}()$ \COMMENT{without closing edge}
		\STATE $s_{\text{best}} \gets s$
		\STATE $\text{tabuList} \gets \text{empty queue}$
		\FOR{$\text{iteration} = 1$ \TO $\text{maxIterations}$}
			\STATE $\text{bestNeighbor} \gets \text{None}$
			\STATE $\text{bestMove} \gets \text{None}$
			\FOR{each $2$-Opt move $(i,j)$ in the current tour}
				\STATE $s' \gets \text{tour after applying move }(i,j)$
				\STATE $\text{isTabu} \gets (i,j) \in \text{tabuList}$
				\IF{$\text{not isTabu}$ \OR $\text{Cost}(s') < \text{Cost}(s_{\text{best}})$}
					\IF{$\text{Cost}(s') < \text{Cost}(\text{bestNeighbor})$}
						\STATE $\text{bestNeighbor} \gets s'$
						\STATE $\text{bestMove} \gets (i,j)$
					\ENDIF
				\ENDIF
			\ENDFOR
			\IF{$\text{bestNeighbor} = \text{None}$} \STATE \textbf{break} \ENDIF
			\STATE $s \gets \text{bestNeighbor}$
			\STATE add $\text{bestMove}$ to $\text{tabuList}$
			\IF{$\text{len}(\text{tabuList}) > \text{tenure}$}
				\STATE remove oldest entry
			\ENDIF
			\IF{$\text{Cost}(s) < \text{Cost}(s_{\text{best}})$}
				\STATE $s_{\text{best}} \gets s$
			\ENDIF
		\ENDFOR
		\RETURN $s_{\text{best}}$ with closing edge
	\end{algorithmic}
\end{algorithm}

The selected move is then applied to become the new current tour, and the move (represented as the pair $(i,j)$) is added to the tabu list.
If the tabu list exceeds a given \emph{tabu tenure}, the oldest entry is removed.
The tabu tenure is set to $\max(7, \lfloor\sqrt{n}\rfloor)$.
This choice follows a common static rule depending on the problem size, where $\sqrt(n)$ 
provides a balanced trade-off between intensification (not overly restricting the search) 
and diversification (preventing cycles) \cite{krishnakumarriyerTabuSearchAlgorithm2013}  \textcolor{red}{cited this, but it's behind a paywall}\cite{osmanMetastrategySimulatedAnnealing1993}

The search continues for a fixed number of iterations, which depends on the problem size: \newline
$\text{MAX\_ITERATIONS} = \max(50,\, 5000 / n)$

This scaling ensures that large instances are not overwhelmed by an excessive number of iterations, while small instances still receive enough exploration.
The algorithm returns the best tour found over the whole search.

The aspiration criterion used here is the simplest and most common one.
A tabu move is allowed if it produces a tour strictly shorter than the global best tour found so far \cite[Sec. 2.3.5]{gendreauTabuSearch2019}.
This guarantees that no improving move that would set a new record is ever rejected.
The tabu list stores only the pair $(i,j)$ representing the inversion endpoints.
This prevents the immediate reversal of that particular 2-Opt swap, 
but does not forbid other moves that involve the same edges or cities, thus keeping the search sufficiently flexible.

Compared to Simulated Annealing, Tabu Search uses a deterministic move selection rule (best admissible neighbor) 
and relies on explicit memory to avoid cycles rather than on probabilistic acceptance.
For the TSP, TS often finds very good solutions in a moderate number of iterations,
because its best-improvement strategy aggressively descends toward local optima,
while the tabu list provides just enough diversification to escape them.
The implementation follows the classic template described in \cite[Sec. 2.3.6]{gendreauTabuSearch2019}.






\subsection{Iterated Local Search}
The idea of iterated local search is that instead of searching the whole space of all candidate solutions, 
we focus only on candidate solutions in the local neighborhood of an initial solution.
These are usually generated by some kind of local search heuristic.
It is conceptually simple, but has led to state-of-the-art algorithms for many problems \cite{lourencoIteratedLocalSearch2019}.
In its essence, iterated local search builds a chain of solutions by exploring the currently best solutions local neighborhood 
but introduces some randomness in order to not get stuck in local minima.

\begin{algorithm}[H]
	\caption{Iterated Local Search}
	\label{alg:ils}
	\begin{algorithmic}[1]
		\STATE $s_0 = \text{GenerateInitialSolution}()$
		\STATE $s^* = \text{LocalSearch}(s_0)$
		\STATE $s_{\text{best}} = s^*$
		\REPEAT
		\STATE $s' = \text{Perturbation}(s^*)$
		\STATE $s^{*\prime} = \text{LocalSearch}(s')$
		\IF{$\text{Cost}(s^{*\prime}) < \text{Cost}(s_{\text{best}})$}
		\STATE $s_{\text{best}} = s^{*\prime}$
		\ENDIF
		\IF{$\text{Cost}(s^{*\prime}) < \text{Cost}(s^*)$}
		\STATE $s^* = s^{*\prime}$
		\ENDIF
		\UNTIL{termination condition met (e.g., $\text{iterations} = 200$)}
		\RETURN $s_{\text{best}}$
	\end{algorithmic}
\end{algorithm}
Iterated local search can be used with any heuristic optimization algorithm (\textit{LocalSearch}) that takes in a tour and returns one that is at least as good as the input.
It starts from some kind of initial solution \( s_{0} \) that is improved to \( s^* \) by one iteration of the \textit{LocalSearch} algorithm. This is the current solution.
Then we try to improve \( s^* \) for a number of iterations or until a termination condition is met.

For the algorithm to be able to escape the local neighborhood, instead of applying \textit{LocalSearch} directly to the current solution \( s^* \) again,
first a change or \textit{Perturbation} is applied that leads to an intermediate solution \( s' \).
Then \textit{LocalSearch} is applied to \( s' \) and we get a new candidate solution \( s^*\prime ' \).
If this candidate solution \( s^*\prime ' \) has a better \textit{Cost} than the current one, it becomes the new current solution \( s^*\prime \).
If it is better than the current best, it also becomes the new current best \( s_{best} \).


The \textit{Cost} is the function that is to be minimized by the algorithm.
In the case of the TSP that is the length of the tour.

The \textit{Perturbation} is the mechanism that allows the algorithm to escape local minima. This function should not be reversible by the \textit{LocalSearch}.
So if the algorithm steps from \( s^*_1 \to s^*_2 \) it then should be unable to do \(  s_2^* \to  s_1^* \) in the same iteration.

Usual choices, and what was also used in this study are 2-Opt for the \textit{LocalSearch} and 4-Opt for the \textit{Perturbation},
this is a good choice because a 4-Opt move cannot be undone by a 2-Opt move and thereby return to the last starting point of the current iteration.

4-Opt (also called double-bridge move) randomly splits the tour into four segments \( A+B+C+D \) and reconnects them as \( A+C+B+D \) \cite{lourencoIteratedLocalSearch2019}.






\subsection{Genetic Algorithm}
Genetic Algorithms (GAs) are population-based metaheuristics inspired by natural evolution \cite{whitleyNextGenerationGenetic2019}.
A population of candidate tours evolves over several generations using selection, crossover, and mutation.
The algorithm maintains a diverse set of solutions, which helps to explore the search space more broadly than local search alone.

\subsubsection{Representation and Initialization}
Each tour is represented as a permutation of the \(n\) city indices (without the closing edge).
The initial population consists of \( \text{POPULATION\_SIZE} = 100 \) individuals.
One individual is created by the same greedy nearest-neighbor heuristic used for the other algorithms.
The remaining 99 individuals are random permutations of the cities.
This hybrid initialization ensures that a good starting solution is always present while providing diversity.

\begin{algorithm}[H]
	\caption{Genetic: Initialization of the Population}
	\label{alg:ga-init}
	\begin{algorithmic}[1]
		\STATE $n \gets \text{number of cities}$
		\STATE $\text{popSize} \gets 100$
		\STATE $\text{population} \gets \text{empty list}$
		\STATE $\text{greedy} \gets \text{GreedyInitialTour}()$ \COMMENT{without closing edge}
		\STATE $\text{population.append}(\text{greedy})$
		\WHILE{$\text{len}(\text{population}) < \text{popSize}$}
			\STATE $\text{randomTour} \gets \text{random permutation of } \{0,\dots,n-1\}$
			\STATE $\text{population.append}(\text{randomTour})$
		\ENDWHILE
		\RETURN $\text{population}$
	\end{algorithmic}
\end{algorithm}

\subsubsection{Selection and Elitism}
At each generation, the next population is built as follows.
First, the \(\text{ELITISM\_COUNT} = 2\) best tours of the current population are copied unchanged into the new population (\emph{elitism}), 
guaranteeing that the best solution never degrades.
The remaining slots are filled by repeatedly selecting two parents using \emph{tournament selection}:
Pick \(\text{TOURNAMENT\_SIZE} = 5\) random individuals from the population and return the one with the shortest tour length.
This is done twice to obtain two parents.
Tournament selection applies selection pressure without requiring global sorting.

\begin{algorithm}[H]
	\caption{Genetic: Tournament Selection}
	\label{alg:ga-select}
	\begin{algorithmic}[1]
		\REQUIRE population, costs, tournamentSize
		\STATE $\text{indices} \gets \text{random choice of tournamentSize}$ \\
		\STATE $\hspace{10em} \mbox{\textcolor{red}{$\hookrightarrow$}} \text{indices from population}$
		\STATE $\text{bestIdx} \gets \text{index in indices with minimum cost}$
		\RETURN $\text{population}[\text{bestIdx}]$
	\end{algorithmic}
\end{algorithm}

\subsubsection{Crossover and Mutation}
With probability \(\text{CROSSOVER\_RATE} = 0.9\), the two parents are recombined using \emph{Order Crossover (OX)}.
OX works by:
\begin{enumerate}
    \item choosing two random cut points \(i < j\) in the parent permutation,
    \item copying the segment between \(i\) and \(j\) from the first parent into the child at the same positions,
    \item filling the remaining positions in the child with the cities from the second parent in the order they appear, skipping those already placed.
\end{enumerate}

\begin{algorithm}[H]
	\caption{Genetic: Order Crossover (OX)}
	\label{alg:ga-ox}
	\begin{algorithmic}[1]
		\REQUIRE parent1, parent2 (permutations of $0,\dots,n-1$)
		\STATE $n \gets \text{len}(\text{parent1})$
		\STATE $i,j \gets$ uniformly random positions with $i \le j$
		\STATE $\text{child} \gets [\;]$ with length $n$
		\STATE $\text{child}[i:j] \gets \text{parent1}[i:j]$
		\STATE $\text{fill} \gets$ cities in parent2 not in $\text{child}[i:j]$
		\STATE $k \gets n - j - 1$   \COMMENT{number of positions after $j$}
		\STATE child$[j + 1: n] \gets$ fill$[0: k]$ preserving cyclic order
		\STATE child$[0:i-1] \gets$ fill$[k+1:]$
		\RETURN child
	\end{algorithmic}
\end{algorithm}



This operator preserves the relative order of cities from both parents and produces a valid tour.
If crossover is not applied (with probability \(0.1\)), the child is an exact copy of the first parent.

After crossover, each child is mutated with probability \(\text{MUTATION\_RATE} = 1/n\).
The mutation operator is \emph{inversion}: Two random positions \(i < j\) are chosen, and the segment between them is reversed.
Inversion is a common mutation for permutation representations because it preserves the set of edges (i.e., does not introduce duplicate cities).

\subsubsection{Generations and Termination}
The number of generations is set to \(\max(100,\, 10 \cdot n)\), where \(n\) is the number of cities.
This scaling gives more generations to larger problems, allowing the GA to converge.
After each generation, the best tour in the new population is compared with the global best found so far.
The global best is updated if an improvement is found.

The algorithm returns the global best tour after the final generation.
\begin{algorithm}[H]
	\caption{Genetic: Inversion Mutation}
	\label{alg:ga-mutate}
	\begin{algorithmic}[1]
		\REQUIRE tour (list of cities)
		\STATE $n \gets \text{len}(\text{tour})$
		\STATE $i,j \gets \text{two distinct random positions in } [0,n-1]$, with $i \le j$
		\STATE $\text{tour}[i:j+1] \gets \text{reversed}(\text{tour}[i:j+1])$
		\RETURN $\text{tour}$
	\end{algorithmic}
\end{algorithm}



The genetic algorithm does not use any local improvement.
It relies entirely on crossover and mutation to evolve good tours.
Despite this, the combination of a greedy initial solution, elitism, 
and the order crossover operator often yields competitive results, especially for smaller instances.
For larger problems, the GA can be slower than local search methods, 
but it maintains a broad exploration of the search space.


\begin{algorithm}[H]
	\caption{Genetic Algorithm}
	\label{alg:ga}
	\begin{algorithmic}[1]
		\STATE $\text{population} \gets \text{InitializePopulation}()$ \COMMENT{1 greedy + 99 random}
		\STATE $s_{\text{best}} \gets \text{best tour in population}$
		\FOR{$\text{generation} = 1$ \TO $\text{maxGenerations}$}
			\STATE $\text{newPopulation} \gets \text{elite individuals from population}$
			\WHILE{$\text{newPopulation not full}$}
				\STATE $p_1 \gets \text{TournamentSelect}(\text{population})$
				\STATE $p_2 \gets \text{TournamentSelect}(\text{population})$
				\IF{random number $< \text{crossoverRate}$}
					\STATE $\text{child} \gets \text{OrderCrossover}(p_1, p_2)$
				\ELSE
					\STATE $\text{child} \gets p_1$
				\ENDIF
				\IF{random number $< \text{mutationRate}$}
					\STATE $\text{child} \gets \text{InversionMutate}(\text{child})$
				\ENDIF
				\STATE add $\text{child}$ to $\text{newPopulation}$
			\ENDWHILE
			\STATE $\text{population} \gets \text{newPopulation}$
			\STATE $s_{\text{best}} \gets \min(s_{\text{best}},\; \text{best in population})$
		\ENDFOR
		\RETURN $s_{\text{best}}$ with closing edge
	\end{algorithmic}
\end{algorithm}

An advanced discussion of genetic algorithms for combinatorial optimization, 
including Gray-box optimization and the EAX crossover, is given in \cite{whitleyNextGenerationGenetic2019}.