\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
%Template version as of 6/27/2024

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}



\begin{document}

\title{Conference Paper Title*\\
{\footnotesize \textsuperscript{*}Note: Sub-titles are not captured for https://ieeexplore.ieee.org  and
should not be used}
\thanks{Identify applicable funding agency here. If none, delete this.}
}

\author{\IEEEauthorblockN{1\textsuperscript{st} Daniel Schafhäutle}
\IEEEauthorblockA{\textit{Computational and Data Science} \\
\textit{FHGR}\\
Chur, Switzerland \\
daniel.schafhautle@stud.fhgr.ch}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Joshua Kohler}
\IEEEauthorblockA{\textit{Computational and Data Science} \\
\textit{FHGR}\\
Chur, Switzerland \\
joshua.kohler@stud.fhgr.ch}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Gian-Luca Lanfranchi}
\IEEEauthorblockA{\textit{Computational and Data Science} \\
\textit{FHGR}\\
Chur, Switzerland \\
gianluca.lanfranchi@stud.fhgr.ch}
}

\maketitle

\input{content/2_abstract.tex}
\input{content/3_intro.tex}
\input{content/4_relatedwork.tex}
\input{content/5_materialsandmethods.tex}
\input{content/6_experimentsandresults.tex}
\input{content/7_discussion.tex}
\input{content/8_conclusion.tex}


\textcolor{red}{Wollt ihr die Referenzen manuell oder mit z.B. Zotero machen?}
\begin{thebibliography}{00}
    \bibitem{b1} G. Eason, B. Noble, and I. N. Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil. Trans. Roy. Soc. London, vol. A247, pp. 529--551, April 1955.
\end{thebibliography}


\input{content/9_appendix.tex}

\end{document}
